A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-24 jacobian_calc_CMAME.tex
More edits

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%%\usepackage{graphics}
%% or use the graphicx package for more complicated commands
%%\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
%%\usepackage{color}
%%\usepackage{tkz-base}
\usepackage{pgfplots}
\pgfplotsset{compat=1.8}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{array}
\usepackage{hyperref}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}

%Package for adding notes/comments/and tracking changes
\usepackage[]{changes}
\usepackage[]{todonotes}

@@

    In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the \todo[inline]{Re: I.1} method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was \todo[inline]{Re: I.1} implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\todo[disable]{Re: II.1}\footnote{In the context of plasticity thoery, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \todo[disable]{Re: I.1} relatively unexplored, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This  \todo[disable]{Re: I.1} method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

@@

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the  \todo[disable]{Re: I.1} complex-step method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward-difference, central-difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by the study.

The aim of this paper is not only to \todo[disable]{Re: I.1} study the complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the  \todo[disable]{Re: I.1} complex-step method for calculating tangent-stiffness matrices, in Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software, in Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods, in Section~\ref{sec:Res} a presentation and analysis of the results of the comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion.  \todo[disable]{Re: II.5}This open source resource is available at the corresponding author's website or at \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left(h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\todo[disable]{Re: II.6}For completeness, the forward- and central-difference formulas corresponding to equation~(\ref{eqn:complexFirstDeriv}) are \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
The order of the truncation error in the forward-difference formula (equation~(\ref{eqn:forwardFirstDeriv})) indicates that for a given $h$ this method is less accurate an approximation than either central-difference or complex-step which have errors of $\mathcal{O}(h^2)$.

\todo[disable]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central-difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

@@

\todo[disable][]{Re: I.5}
\section{Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:WorkingExamples}

In order to illustrate a comparison of the numerical differentiation techniques in a simpler form than it appears in the large-scale analysis code covered in the following sections, a simple example borrowed from \cite{rezaiee2010dynamic} was implemented with each of the methods, CS, AD, and FD. All the source code to reproduce the results described here are available in the associated paper repository.

\subsection{Common code} 
\label{subsec:CommonCode}

The source code for the header-only library is entitled \emph{NewtonRaphson.hpp}. Note that the library is an interface to functions and data structures made available in the \emph{Trilinos} library, mentioned earlier, such that it is required in order for the examples described here to work. \emph{Trilinos} is freely available for download as of this writing at: \emph{trilinos.sandia.gov}.  Generally the header-only library works by taking in the users initial guessed equilibrium solution, target dependent variable values, system constants, and a function pointer to a model evaluation function. Solver parameters including an update scaling factor, tolerance, maximum iterations and probe length are chosen after the solver is instantiated as a new object. The behavior of the solver is specialized through a coding feature called inheritance to perform each of CS, AD and FD.  But, in common, each version of the solver works by computing the system Jacobian, solving a linear system as in that step of the Newton Raphson method, computing the residual and checking for convergence or overrun. Here the linear solution method is LU decomposition, while the chief convergence criterion is change in residual between iterations.  Additionally the problem can be partially respecified to allow for running a series of load steps having progressing target values, which is a typical mode also seen in \emph{Peridigm}. Applying boundary conditions incrementally in load steps assists convergence.

\subsection{Second example and related studies} 
\label{subsec:Validation}
The second program,a \emph{NRComparison.cpp}, follows a case study that appears in \cite{rezaiee2010dynamic}. The example program departs from the cited literature by applying each of CS, AD and FD and verifying them against an analytical model of a nonlinear mechanical system involving a connected beam and spring. Fig. \ref{fig:TrussSchematic} is a schematic of the beam and spring system. 
%
\begin{figure}[tbp] 
\centering %%\includegraphics[width=0.7\textwidth]{./figs/truss.png}

\begin{tikzpicture}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{calc}
\usetikzlibrary{patterns}

\tikzstyle{wall}=[fill,pattern=north east lines,draw=none,minimum width=0.2cm,minimum height=5.1cm];
\tikzstyle{ground}=[fill,pattern=north east lines,draw=none,minimum width=6.1cm,minimum height=.2cm];

\coordinate (bar_1) at (0,0);
\coordinate (bar_1_support) at (0,-.5);

\coordinate (bar_2) at (5,3);
\coordinate (bar_2_pin) at (5.4,3.0);
\coordinate (spring_1_support) at (5,-.5);
\coordinate (spring_1_pin) at (5,0);

\coordinate (rhm) at (5.4,1.5);
\coordinate (rhml) at (4.5,1.5);
\coordinate (mf) at (2.5, -0.8);
\coordinate (lm) at (1.0, 0.0);
\coordinate (mm) at (1.2, 0.0);
\coordinate (rht) at (5.0, 4.0);
\coordinate (rhtt) at (5.0, 4.2);

\node (wall_slider) at (bar_2_pin) [circular sector, draw, rotate=180] {};
\node (ground_triangle_1) at (bar_1_support) [isosceles triangle, draw, rotate=90] {};
\node (ground_triangle_2) at (spring_1_support) [isosceles triangle, draw, rotate=90] {};

\node (wall) at ($ (rhm) !.3cm!(10, 3) $)  [rectangle, draw, style=wall] {};
\node (ground) at (mf)  [rectangle, draw, style=ground] {};

\node (spring_label) at (rhml) [] {$k_s$};
\node (load_label) at (rhtt) [] {$P$};
\node (angle_label) at ($ (mm)!.4cm!(bar_2) $) [] {$\phi$};

\path (bar_1) -- (bar_2) [draw, very thick];
\path (bar_1) circle (.10) [fill=black];
\path (bar_2) circle (.10) [fill=black];
\path (spring_1_pin) circle (.10) [fill=black];
\path [decorate, decoration={coil, segment length = 4, amplitude=4}] (bar_2) -- (spring_1_pin) [draw, thin, stroke=black];
\path (rht) -- ($ (bar_2)!.2cm!(rht) $) [draw, -latex, ultra thick, color=black];
\path (lm) arc(0:16:2) (lm) [draw];
\path (bar_1) -- (mm) [draw, -latex];
--cycle
\end{tikzpicture}

\caption{Beam and spring simulated in example program.} \label{fig:TrussSchematic}
\end{figure}
%
The equations describing this system and results for verification were obtained
in \cite{rezaiee2010dynamic}. They include a force function, Eqn.
\ref{eqn:TrussForce}, and an analytical force derivative or stiffness, Eqn.
\ref{eqn:TrussStiffness}.

\begin{equation} 
    \label{eqn:TrussForce}
    f(D) = .5AE(\cos^{2}\phi)(\frac{D}{L_{0}})^{2}[\frac{D}{L_{0}}\cos^{2}\phi - 3\sin\phi] + k_{s}D + (AE\frac{D}{L_{0}})\sin^{2}\phi,
\end{equation}

\begin{equation} 
    \label{eqn:TrussStiffness}
    S_{T}(D) = 1.5AE(\cos^{2}\phi)[\frac{D}{L_{0}}\cos^{2}\phi - 2sin\phi](\frac{D}{L_{0}^{2}}) + k_{s} + \frac{AE\sin^{2}\phi}{L_{0}}.
\end{equation}

Where $\phi$ is a constant, beam angle, $AE$ is rigidity, $D$ is displacement,
$L_{0}$ is original beam length, and $k_{s}$ is the Hooke's constant for the
coil spring attached to the beam. For a verification of each of the methods as
well as a convergence rate study, this truss and spring model was analyzed in
the second program. $f(D)$ is restorative force to balance downward force
loading at the connection point of the beam and the spring, parallel to the
wall. As in the reference, equilibrium solutions for 12 incremental load steps
of 4.448 Newtons were sought. Differing from the reference, Newton's method was
used rather than 'Dynamic Relaxation'. Solver parameters common to all tests
were a limit of 100 iterations for convergence, a residual change tolerance of
$1\times{10^{-9}} \si{\pico\meter}$ and an update multiplier of 1.0. Data
collected included number of iterations to converge summed over all load-steps
and final displacement, D, after the final load-step. Additionally
finite difference step-size was varied in order to measure the impact on each
of the Jacobian calculation methods. Naturally, varying step size does not
affect AD or the analytically derived Jacobian since those are not functions
of step size. The accuracy results for
verifying the Jacobian calculation methods are summarized in Table
\ref{tab:ConvergenceStudy}.

\begin{table*}[!ht]    
  \centering
        \caption{Displacement, D, after load-step 12} \label{tab:Verification}   
        \begin{tabular}{c c c c c}
         \toprule
         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 
        \midrule
        0 & 5.07996 & " & " & "\\
        -4 & 5.07996 & " & " & "\\
        -8 & 5.07996 & " & " & "\\
        -12 & 5.07996 & " & " & " \\
        -16 & 5.07996 & " & " & failed \\
        -20 & 5.07996 & " & " & failed \\
        \bottomrule
    \end{tabular}
\end{table*}

Table \ref{tab:Verification} shows that each of the methods solved the problem
as accurately as one another within the precision displayed on the table. The
exception to this is that the finite difference method failed due to an
arithmetic error in association with extremely small step size, suggesting
subtractive cancellation. However, since there is no accuracy reason for
choosing an extremely small step size, each of the methods can be used to
produce acceptable solutions to the nonlinear truss system. Table
\ref{tab:ConvergenceStudy} shows the total number of iterations taken over all
load-steps for each of the methods.

>>>>

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%%\usepackage{graphics}
%% or use the graphicx package for more complicated commands
%%\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
%%\usepackage{color}
%%\usepackage{tkz-base}
\usepackage{pgfplots}
\pgfplotsset{compat=1.8}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{array}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage{upquote}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}

%Package for adding notes/comments/and tracking changes
\usepackage[color=red!40, textsize=scriptsize]{todonotes}

@@

    In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the  method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\todo[]{Re: II.1}\footnote{In the context of plasticity thoery, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \todo[]{Re: I.1} relatively unexplored, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This  \todo[]{Re: I.1} method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

@@

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the  \todo[]{Re: I.1} complex-step method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward difference, central difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by the study.

The aim of this paper is not only to \todo[]{Re: I.1} study the complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the  \todo[]{Re: I.1} complex-step method for calculating tangent-stiffness matrices, in Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software, in Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods, in Section~\ref{sec:Res} a presentation and analysis of the results of the comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion.  \todo[]{Re: II.5}This open source resource is available at the corresponding author's website or at \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left(h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\todo[]{Re: II.6}For completeness, the forward and central difference formulas corresponding to equation~(\ref{eqn:complexFirstDeriv}) are \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
The order of the truncation error in the forward difference formula (equation~(\ref{eqn:forwardFirstDeriv})) indicates that for a given $h$ this method is less accurate an approximation than either central difference or complex-step which have errors of $\mathcal{O}(h^2)$.

\todo[]{Re: I.4}As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler \cite{lyness1967numerical}.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through numerical experiments that CS offers superior accuracy to FD and CD \cite{squire1998using} in computing numerical derivatives. Papers by Newman et al.\ \cite{newman1998} and Anderson et al.\ \cite{anderson2001sensitivity} discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ \cite{perez2000numerical, perez2012numerical} used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ \cite{martins2000automated} investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.

@@

\section{\todo[]{Re: I.5}Simple example of CS, AD and FD for solution of nonlinear problems.}
\label{sec:WorkingExamples}

In order to illustrate a comparison of the numerical differentiation techniques in a simpler form than it appears in the large-scale analysis code covered in the following sections, a simple example borrowed from \cite{rezaiee2010dynamic} was implemented with each of the methods, CS, AD, and FD. All the source code to reproduce the results described here are available in the associated paper repository. This source describes the implemenation and interface to the \emph{Trilinos Secado library} for automatic differentiation.

The example problem is a single degree-of-freedom nonlinear mechanical system involving a connected truss member and spring as illustrated in Fig. \ref{fig:TrussSchematic}. 
%
\begin{figure}[tbp] 
\centering %%\includegraphics[width=0.7\textwidth]{./figs/truss.png}

\begin{tikzpicture}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{decorations.pathmorphing}
\usetikzlibrary{calc}
\usetikzlibrary{patterns}
\tikzstyle{wall}=[fill,pattern=north east lines,draw=none,minimum width=0.2cm,minimum height=5.1cm];
\tikzstyle{ground}=[fill,pattern=north east lines,draw=none,minimum width=6.1cm,minimum height=.2cm];
\coordinate (bar_1) at (0,0);
\coordinate (bar_1_support) at (0,-.5);
\coordinate (bar_2) at (5,3);
\coordinate (bar_2_pin) at (5.4,3.0);
\coordinate (spring_1_support) at (5,-.5);
\coordinate (spring_1_pin) at (5,0);
\coordinate (rhm) at (5.4,1.5);
\coordinate (rhml) at (4.5,1.5);
\coordinate (mf) at (2.5, -0.8);
\coordinate (lm) at (1.0, 0.0);
\coordinate (mm) at (1.2, 0.0);
\coordinate (rht) at (5.0, 4.0);
\coordinate (rhtt) at (5.0, 4.2);
\node (wall_slider) at (bar_2_pin) [circular sector, draw, rotate=180] {};
\node (ground_triangle_1) at (bar_1_support) [isosceles triangle, draw, rotate=90] {};
\node (ground_triangle_2) at (spring_1_support) [isosceles triangle, draw, rotate=90] {};
\node (wall) at ($ (rhm) !.3cm!(10, 3) $)  [rectangle, draw, style=wall] {};
\node (ground) at (mf)  [rectangle, draw, style=ground] {};
\node (spring_label) at (rhml) [] {$k_s$};
\node (load_label) at (rhtt) [] {$P$};
\node (angle_label) at ($ (mm)!.4cm!(bar_2) $) [] {$\phi$};
\path (bar_1) -- (bar_2) [draw, very thick];
\path (bar_1) circle (.10) [fill=black];
\path (bar_2) circle (.10) [fill=black];
\path (spring_1_pin) circle (.10) [fill=black];
\path [decorate, decoration={coil, segment length = 4, amplitude=4}] (bar_2) -- (spring_1_pin) [draw, thin, stroke=black];
\path (rht) -- ($ (bar_2)!.2cm!(rht) $) [draw, -latex, ultra thick, color=black];
\path (lm) arc(0:16:2) (lm) [draw];
\path (bar_1) -- (mm) [draw, -latex];
--cycle
\end{tikzpicture}
\caption{Beam and spring simulated in example program.} 
\label{fig:TrussSchematic}
\end{figure}
%
For this system, the equalibrium reaction force associated with a displacement $D$ of the spring is shown in equation~(\ref{eqn:TrussForce}) 
%
\begin{equation} 
    \label{eqn:TrussForce}
    f(D) = \frac 1 2 AE(\cos^{2}\phi)(\frac{D}{L_{0}})^{2}[\frac{D}{L_{0}}\cos^{2}\phi - 3\sin\phi] + k_{s}D + (AE\frac{D}{L_{0}})\sin^{2}\phi,
\end{equation} 
%
and the analytical tangent-stiffness is given by equation~(\ref{eqn:TrussStiffness})
%
\begin{equation} 
    \label{eqn:TrussStiffness}
    S_{T}(D) = \frac 3 2 AE(\cos^{2}\phi)[\frac{D}{L_{0}}\cos^{2}\phi - 2sin\phi](\frac{D}{L_{0}^{2}}) + k_{s} + \frac{AE\sin^{2}\phi}{L_{0}}.
\end{equation}
%
where $\phi$ is a constant truss angle, $AE$ is the axial rigidity of the truss, $L_{0}$ is original truss member length, and $k_{s}$ is the Hooke's constant for the spring attached to the truss member. For a demonstration and verification of each of the methods as well as a convergence rate study, this truss and spring model was analyzed with each of the methods and compared with the analytic solution. As in the reference \cite{rezaiee2010dynamic}, equilibrium solutions for 12 incremental load steps of $\Delta P=\SI{4.448}{\newton}$ were found. Solver parameters common to all tests were a maximum limit of 100 iterations for convergence, a residual change tolerance of $\SI{1e-9}{\meter}$.  Data collected included number of iterations to converge summed over all load-steps and final displacement $D$, after the final load-step. Additionally finite difference step-size was varied in order to measure the impact on each of the approximate tangent-stiffness calculation methods. Naturally, varying step size does not affect AD or the analytically derived tangent-stiffness since those are not functions of step size. The accuracy results for verifying the tangent-stiffness calculation methods are summarized in Table~\ref{tab:ConvergenceStudy}. 
%
\begin{table}[tbp]    
  \centering
        \caption{Displacement, D, after load-step 12} \label{tab:Verification}   
        \begin{tabular}{c c c c c}
         \toprule
         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 
        \midrule
        0 & 5.07996 & " & " & "\\
        -4 & 5.07996 & " & " & "\\
        -8 & 5.07996 & " & " & "\\
        -12 & 5.07996 & " & " & " \\
        -16 & 5.07996 & " & " & failed \\
        -20 & 5.07996 & " & " & failed \\
        \bottomrule
    \end{tabular}
\end{table}
%
Table.~\ref{tab:Verification} shows that each of the methods solved the problem as accurately as one another within the precision reported. The exception to this is that the finite difference method failed due to an arithmetic error in association with extremely small step size, suggesting subtractive cancellation. However, since there is no accuracy reason for choosing an extremely small step size, each of the methods can be used to produce acceptable solutions to the nonlinear truss system. Table~\ref{tab:ConvergenceStudy} shows the total number of iterations taken over all load-steps for each of the methods. 
%
\begin{table}[tbp]    
  \centering
        \caption{Summed iterations over all load steps} \label{tab:ConvergenceStudy}   
        \begin{tabular}{c c c c c}
         \toprule
         Step exponent & Actual $\si{\centi\meter}$ & AD $\si{\centi\meter}$ & CS $\si{\centi\meter}$ & FD $\si{\centi\meter}$\\ 
        \midrule
        0 & 63 & " & 213 & 417\\
        -4 & 63 & " & " & "\\
        -8 & 63& " & " & "\\
        -12 &63& " & " &  65\\
        -16 &63& " & " & fail\\
        -20 &63& " & " & fail\\
        \bottomrule
    \end{tabular}
\end{table}
%
Table~\ref{tab:ConvergenceStudy} indicates that when using AD or the `actual' analytically derived tangent-stiffness, a minimum number of iterations is achieved corresponding to a maximum convergence rate. CS and FD are similarly affected by a grossly large step-size, while subtractive cancellation invalidates the results for FD for very small step-sizes. Inaccuracy for large step-sizes is as predicted by the FD and CS formulas, which require a small step-size to allow the neglect of higher order terms. The results of this test show that step-size selection is not unimportant when using FD, but when properly configured any of the methods may return the same maximum convergence rate for this problem.  Overall conclusions regarding the methods that the accuracy of the tangent-stiffness should not be confused with the accuracy of the equilibrium solution, since each of the methods discussed here may be equally capable given enough iterations. Numerical measurements of accuracy besides number of iterations taken are not possible here because there is not basis for comparison, as the methods are solving the problem separately and producing different local Newton iterates. Additionally, while the inaccuracy of the tangent-stiffness may strongly affect convergence rates, this relationship was not precisely quantified here.
